---
title: "chl-mvco-eml-assembly"
author: "Kate Morkeski"
date: "2022-06-14"
output: html_notebook
---

## Setup

Libraries used

```{r}
# two of the required packages are installed from GitHub
# library(remotes)
# remotes::install_github("EDIorg/EMLassemblyline")
# remotes::install_github("WHOIGit/ediutilities")

library(EMLassemblyline)
library(ediutilities)
library(here)
library(readxl)
library(lubridate)
library(tidyverse)
library(httr)
library(compareDF)

```

Read data table

```{r}

chl_input <- read_excel(here('MVCO_chl_4EDI_copy20240617.xlsx')) 

```

Format columns

```{r}

names(chl_input) <- tolower(names(chl_input))

# combine date and time
chl_input$start_date <- as.character(chl_input$start_date) 
chl_input$start_time_utc <- as.character(chl_input$start_time_utc)
chl_input$start_time_utc <- gsub("1899-12-31 ", "", chl_input$start_time_utc)
chl_input$date_time_utc <- paste(chl_input$start_date, chl_input$start_time_utc)
chl_input$date_time_utc <- as.POSIXct(chl_input$date_time_utc, tz = "GMT", format="%Y-%m-%d %H:%M:%OS")
# does formatting as GMT have any consequences compared to actual UTC?

# format columns for new dataframe
chl_mvco <- chl_input %>%
  rename(chl = 'chl (ug/l)') %>%
  rename(phaeo = 'phaeo (ug/l)') %>%
  select(-start_date, -start_time_utc)  %>%
  relocate(date_time_utc, .before = latitude) %>%
  relocate(filter_size, .before = replicate) %>%
  mutate(latitude = round(latitude, 4)) %>% 
  mutate(longitude = round(longitude, 4)) %>% 
  mutate(chl = round(chl, 3)) %>%
  mutate(phaeo = round(phaeo, 3))

# edit filter size designation to include prefilter
# modify filter size designation for rows with prefilter size in replicate column
chl_mvco <- chl_mvco %>%
  mutate(filter_size = as.character(filter_size)) %>% 
  mutate(filter_size = case_when(filter_size == "0" ~ ">0",
                                 filter_size == "10" ~ ">0&<10",
                                 filter_size == "80" ~ ">0&<80",
                                 TRUE ~ filter_size)) %>% 
  mutate(filter_size = case_when(replicate == "10a" ~ ">0&<10",
                                 replicate == "10b" ~ ">0&<10",
                                 replicate == "10c" ~ ">0&<10",
                                 TRUE ~ filter_size))

# strip filter size out of replicate column
chl_mvco$replicate <- gsub("10", "", chl_mvco$replicate)
chl_mvco$replicate <- gsub("80", "", chl_mvco$replicate)

chl_mvco <- chl_mvco %>% relocate(depth, .before = filter_size)

```

Handle quality flags

```{r}

unique(chl_input$quality_flag)

# convert quality flag from lab standard to IODE 
chl_mvco <- chl_mvco %>%
   mutate(iode_quality_flag = case_when(is.na(chl) ~ 9,
                                        quality_flag == "1" ~ 1,
                                        is.na(quality_flag) ~ 2,
                                        quality_flag == "2" ~ 3,
                                        quality_flag == "3" ~ 4,
                                        ))
# rows with missing chl are retained, with flag indicating missing

# check on distribution of quality flags
hist(chl_mvco$iode_quality_flag)
unique(chl_mvco$iode_quality_flag)
chl_bad <- filter(chl_mvco, iode_quality_flag == 4) # 30 samples have flag 4
low_phaeo <- filter(chl_mvco, phaeo < 0)    # 11 samples have negative phaeo, 10 already flagged 3 or 4

# replace chlorophyll and phaeo with NA for any samples flagged bad data 
# convert quality flag for any samples with negative phaeo
chl_mvco <- chl_mvco %>%
  mutate(chl = replace(chl, iode_quality_flag == "4", NA)) %>%
  mutate(phaeo = replace(phaeo, iode_quality_flag == "4", NA)) %>%
  mutate(iode_quality_flag = replace(iode_quality_flag, phaeo <0 & iode_quality_flag <3, 3))

# update columns
chl_mvco <- chl_mvco %>%
  relocate(iode_quality_flag, .before = fluorometer) %>%
  select(-quality_flag) 
  
```

Add Tioga cruise number

```{r}

df <- read_excel('List of MVCO cruises for NES-LTER project.xlsx', sheet = "catalog", skip = 6)
Tioga_ID <- df %>% select(1, 3, 4, 5)

chl_mvco <- left_join(chl_mvco, Tioga_ID, by = "event_number")

chl_mvco <- chl_mvco %>%
  select(-date)  %>%

  relocate(ship, .before = date_time_utc) %>%
  relocate(cruise_ID, .before = date_time_utc)

chl_mvco <- chl_mvco %>%
  rename(cruise = cruise_ID)  

```

# Merge MVCO samples from transect cruises

```{r}
chl_transect <- read_from_api(type = "chl", "all")

# Transect MVCO ranges lat =  41.3125 - 41.3269, lon = -70.5629 to -70.5857
# MVCO_nut_copy ranges lat = 41.1364 - 41.339, lon = -70.4151 to -70.6839
lat_min <- 41.13
lat_max <- 41.33
lon_min <- -70.68
lon_max <- -70.41

# Filter rows within the specified range of latitude and longitude
filtered_chl <- chl_transect %>%
  filter(latitude >= lat_min & latitude <= lat_max,
         longitude >= lon_min & longitude <= lon_max)

filtered_chl <- filtered_chl %>%
  rename(date_time_utc = date)

# Add ship, event_number, event_number_niskin columns and populate
filtered_chl <- filtered_chl %>%
  mutate(ship = case_when(
    grepl("^AR", cruise) ~ "Armstrong",
    grepl("^AT", cruise) ~ "Atlantis",
    grepl("^EN", cruise) ~ "Endeavor",
    grepl("^HRS", cruise) ~ "Sharp",
    TRUE ~ ""  # Set to empty string for unmatched cases
      ),
    event_number = paste(cruise, cast, sep = "_"),  
    event_number_niskin = paste(cruise, cast, niskin, sep = "_"), 
  )

# Merge transect nut data with nutrients
common_columns <- intersect(names(chl_mvco), names(filtered_chl))
filtered_chl_aligned <- filtered_chl %>% select(all_of(common_columns))
missing_columns <- setdiff(names(chl_mvco), names(filtered_chl_aligned))
filtered_chl_aligned[missing_columns] <- NA
filtered_chl_aligned <- filtered_chl_aligned %>% select(names(chl_mvco))
chl_updated <- bind_rows(chl_mvco, filtered_chl_aligned)

# Round lat/long and chl & phaeo
chl_updated <- chl_updated  %>% 
  mutate(latitude = round(latitude, 4)) %>% 
  mutate(longitude = round(longitude, 4)) %>%
  mutate(chl = round(chl, 3)) %>%
  mutate(phaeo = round(phaeo, 3))

# modify filter size designation less than 10 or 80
chl_updated <- chl_updated %>%
  mutate(filter_size = as.character(filter_size)) %>% 
  mutate(filter_size = case_when(filter_size == "<10" ~ ">0&<10",
                                 filter_size == "<80" ~ ">0&<80",
                                 TRUE ~ filter_size))

# replace chlorophyll and phaeo with NA for any samples flagged bad data 
# convert quality flag for any samples with negative phaeo
chl_updated <- chl_updated %>%
  mutate(chl = replace(chl, iode_quality_flag == "4", NA)) %>%
  mutate(phaeo = replace(phaeo, iode_quality_flag == "4", NA)) %>%
  mutate(iode_quality_flag = replace(iode_quality_flag, phaeo <0 & iode_quality_flag <3, 3))

# Set iode_quality_flag_phosphate to 2 where it is NA
chl_updated <- chl_updated %>%
  mutate(iode_quality_flag = ifelse(is.na(iode_quality_flag), 2, iode_quality_flag))

chl_updated <- chl_updated %>%
  mutate(fluorometer = ifelse(is.na(fluorometer),"Turner Designs Aquafluor Handheld Serial Number 800446", fluorometer))

```

Create package data file

```{r}

chl_updated <- chl_updated %>%
  mutate(
    ship = ifelse(is.na(ship), "", ship),
    cruise = ifelse(is.na(cruise) | cruise == "NA", "", cruise)
  )

# ensure rows are in time order
chl_mvco <- chl_updated[order(chl_updated$date_time_utc),]
  
# write csv file for package
write.csv(chl_mvco, "nes-lter-chl-mvco.csv", na = "NaN", row.names = FALSE)

```
## Quality checks

Summarize and plot chlorophyll values 

```{r}

summary(chl_mvco)

# check on any rows missing filter_size
no_filter <- filter(chl_mvco, is.na(filter_size))

# plot values vs depth
ggplot(chl_mvco, aes(x=chl, y=-depth, color=filter_size))+ geom_point()
ggsave("chl-mvco-values-vs-depth.png")

chl_mvco$iode_quality_flag <- as.factor(chl_mvco$iode_quality_flag)

# plot values vs cruise date
ggplot(chl_mvco, aes(x=date_time_utc, y=chl, color=filter_size))+ geom_point()
ggsave("chl-mvco-filter-size.png")
ggplot(chl_mvco, aes(x=date_time_utc, y=phaeo, color=filter_size))+ geom_point()
ggsave("phaeo-mvco-filter-size.png")
ggplot(chl_mvco, aes(x=date_time_utc, y=chl, color=depth))+ geom_point()
ggsave("chl-mvco-depth.png")
ggplot(chl_mvco, aes(x=date_time_utc, y=phaeo, color=depth))+ geom_point()
ggsave("phaeo-mvco-depth.png")
ggplot(chl_mvco, aes(x=date_time_utc, y=chl, color=iode_quality_flag))+ geom_point()
ggsave("chl-mvco-iode-quality-flag.png")
ggplot(chl_mvco, aes(x=date_time_utc, y=phaeo, color=iode_quality_flag))+ geom_point()
ggsave("phaeo-mvco-iode-quality-flag.png")

```

Map Sampling Locations

Call the map_locs function from ediutilities package to map the sampling locations. Perform a visual check.

```{r}

# Map Check
map_locs(df = chl_mvco, xvar = "longitude", yvar = "latitude",
         region = "transect", colorvar = "filter_size")

map_locs(df = chl_mvco, xvar = "longitude", yvar = "latitude",
         region = "transect", colorvar = "depth")

```

## Read in previous version MVCO nutrient package to compare to new data set

```{r}

url <- ('https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-nes.21.1&entityid=2a7f068e3256f6453e1b31fc247f5c6f')

httr::GET(url, write_disk(tf <- tempfile(fileext = ".csv")))
tf

chl_mvco_pubd <- read_csv(tf, TRUE, show_col_types = FALSE) 

comparison <- compare_df(chl_mvco, chl_mvco_pubd, c("event_number", "depth"))

create_output_table(comparison, output_type='xlsx', file_name='version_comparison.xlsx')

```

## EML Assembly

Read data table

```{r}

data_table <- read.csv(here('nes-lter-chl-mvco.csv'))

# identify date column for ediutilities package to generate temporal coverage
data_table$date = ymd_hms(data_table$date_time_utc)

```

Read the Excel metadata template and generate text templates used by
EMLassemblyline

```{r}

excel_to_template(here('nes-lter-chl-mvco-info'), 'nes-lter-chl-mvco', rights='CCBY')

```
Generate the package and insert the parent project node into the resulting EML

```{r}

#define package ID
pkg_id <- 'knb-lter-nes.21.2'

make_eml(here(),
         dataset.title='Size-fractionated chlorophyll from the Martha’s Vineyard Coastal Observatory (MVCO), ongoing since 2003 (NES-LTER since 2017)',
         data.table='nes-lter-chl-mvco.csv',
         data.table.description="Chlorophyll from water column bottle samples taken near the Martha's Vineyard Coastal Observatory",
         data.table.name = 'nes-lter-chl-mvco',
         temporal.coverage = temporal_coverage(data_table$date),
         geographic.description = "Martha's Vineyard Coastal Observatory",
         geographic.coordinates = geographic_coordinates(data_table$latitude, data_table$longitude),
         maintenance.description = "ongoing",
         user.id = "NES",
         user.domain = "LTER",
         package.id = pkg_id)

project_insert(pkg_id, filename = 'parent_project_NESI-II_RAPID.txt')


# TODO add additional info with multiple projects
```
